A saturation model for impaired learning with enhanced plasticity: Supplementary material

### 2013-02-26 Notes/MHCsupp.tex
improved references to figs

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:multiWTnopre}
  \Phi = -\frac{2\Delta fq}{M},
\end{equation}
%
where it's worth remembering that $\Delta f<0$.
With pre-training:
%
\begin{equation}\label{eq:multiWTpre}
\begin{aligned}
  \Phi &= 16(\Delta f)^2q \, \frac{(1-2\Delta f)^{M/2-1} (1+2\Delta f)^{M/2-1}}
          {(1-2\Delta f)^M - (1+2\Delta f)^M} \\
       &= -\frac{4\Delta fq}{M} + \CO(\Delta f)^2.
\end{aligned}
\end{equation}
%
So, we see that pre-training will speed up learning when $\Delta f$ is small (in absolute value), as seen in \autoref{fig:multistate_weak}\ref{fig:multistate_weak_learnS}.
On the other hand, if $\Delta f$ is close to $-\half$, pre-training will initially slow down learning a lot due to the factor of $(1+2\Delta f)^{M/2-1}$, as seen in \autoref{fig:multistate_strong}\ref{fig:multistate_strong_learnS}.

Intuitively, the flux depends on the slope of the distribution at the centre of the chain (with an offset due to the difference between $f\pot$ and $f\dep$).
Pre-training has two effects: it produces a slope in the right direction (see \autoref{fig:multistate_weak}\ref{fig:multistate_weak_eq}), but it also reduces the distribution at the centre (see \autoref{fig:multistate_strong}\ref{fig:multistate_strong_eq}).
For small $\Delta f$, the first effect is stronger and learning speeds up.
For larger $\Delta f$, the second effect wins and learning slows down.

@@

There are two effects here as well.
Smaller $\beta$ will increase the probability of crossing the centre of the chain, speeding up learning, but it will also concentrate probability at the ends of the chain, depleting the centre and slowing down learning.
The first effect goes like $1/\beta$, whereas the second goes like $\beta^{M/2}$ and will be more significant for smaller $\beta$ or in a longer chain.

@@

However, this is only the initial, instantaneous rate of change.
Let's look at \autoref{fig:multistate_strong}\ref{fig:multistate_strong_learnS}, for which $\Delta f < \Delta f^*(1,M)$, so that pre-training initially slows down learning for both wild-type and mutant.
We see that the pre-trained curve can rapidly catch up with the un-pre-trained one, due to the mass of probability concentrated at the end of the chain (see \autoref{fig:multistate_strong}\ref{fig:multistate_strong_eq}) drifting to the centre.
This happens sooner for the mutant than the wild-type, due to the stronger depressing transitions.
This means that there is an intermediate range of time-scales over which pre-training does slow down learning in the wild-type but speed it up in the mutant, even outside the desired range of values for $\Delta f$.
As we see in \autoref{fig:multistate_med}\ref{fig:multistate_med_learnS}, even when we are in this range, the period of time for which pre-training impairs learning in the wild-type can be very small.

In conclusion, if we choose $\beta<\beta^*(M)$, $\Delta f < \Delta f^*(1,M)$ and look at intermediate time-scales, we will see that the mutant learns slower than wild-type without pre-training, and that pre-training speeds up learning in the mutant but slows it down in the wild-type.
However, in this case pre-training proceeds much faster in the mutant than wild-type (see \autoref{fig:multistate_med}\ref{fig:multistate_med_learn} and \autoref{fig:multistate_strong}\ref{fig:multistate_strong_learn}), which is \emph{not} seen in the experiment.

@@

@@

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:binWTnopre}
  \Phi = -\Delta f q,
\end{equation}
%
where, once again, it's worth remembering that $\Delta f<0$.
With pre-training:
%
\begin{equation}\label{eq:binWTpre}
\begin{aligned}
  \Phi &= -2\Delta f q.
\end{aligned}
\end{equation}
%
So, we see that pre-training will always speed up learning, unlike what is seen in the experiment.
This can be seen in \autoref{fig:binary}\ref{fig:binary_learnS}.

Now, consider the mutant, for which $q\pot=\beta q\dep=q$, $\beta<1$.
Without pre-training:
%
\begin{equation}\label{eq:binKOnopre}
  \Phi = -\frac{2\Delta f q}{1+\beta},
\end{equation}
%
which is always larger that \eqref{eq:binWTnopre}, unlike what is seen in the experiment.
This is also seen in \autoref{fig:binary}\ref{fig:binary_learn},\ref{fig:binary_learnS}.
As discussed below \eqref{eq:multiKOnopre}, the smaller values of has two effects.
Smaller $\beta$ will increase the probability of depression, speeding up learning, but it will also decrease the probability of being ready for depression, slowing down learning.
The second effect is much smaller for the very short chain considered here.

@@

Now, consider the mutant, for which $q\pot=\beta q\dep=q$, $\beta<1$.
Without pre-training:
%
\begin{equation}\label{eq:multiLinKOnopre}
  \Phi = -2\Delta f q\,\frac{1-\beta^{M-1}}{1-\beta^M}.
\end{equation}
%
This decreases monotonically in the interval $\beta\in[0,1]$, therefore this will always be greater than \eqref{eq:multiLinWTnopre}.
This means that the mutant will initially learn faster than the wild-type.
However, as the wild-type will eventually catch up, and this can happen very quickly, as seen in \autoref{fig:multistate_lin}\ref{fig:multistate_lin_learn},\ref{fig:multistate_lin_learnS}

With pre-training:
%
\begin{equation}\label{eq:multiLinKOpre}
\begin{aligned}
  \Phi &= -4\Delta f q \, \frac{(1+2\Delta f)^{M-1} - \beta^{M-1}(1-2\Delta f)^{M-1}}
          {(1+2\Delta f)^M - \beta^M(1-2\Delta f)^M} \\
       &= -4\Delta f q\,\frac{1-\beta^{M-1}}{1-\beta^M} + \CO(\Delta f)^2.
\end{aligned}
\end{equation}
%
The ratio of this to \eqref{eq:multiLinKOnopre} takes its minimum value at the lowest end of the interval $\Delta f \in \brk{-\half,0}$, where it takes the value
%
\begin{equation}\label{eq:multiLinprenopre}
  \frac{\Phi_{\text{with pre}}}{\Phi_{\text{without pre}}} = \frac{1-\beta^M}{\beta-\beta^M}
   \longrightarrow \frac{M}{M-1} \quad \text{as} \quad \beta\to1,
\end{equation}
%
which is always greater than 1.
Therefore, pre-training will always enhance learning, for both mutant and wild-type.

@@

If we compare gain-increase learning in the mutant to wild-type, there are two effects: the increased transition rates speed up learning, but the equilibrium distribution is shifted to the depressed side (compare \autoref{fig:pooled_plenty}\ref{fig:pooled_plenty_eq}: \ref{fig:pooled_plenty_eq_WT} and \ref{fig:pooled_plenty_eq_KO} or \autoref{fig:pooled_scarce}\ref{fig:pooled_scarce_eq}: \ref{fig:pooled_scarce_eq_WT} and \ref{fig:pooled_scarce_eq_KO}), where there are fewer synapses available for depression and resources are depleted.
When resource depletion is less severe, the first effect dominates and the mutant learns faster than wild-type (see \autoref{fig:pooled_plenty}\ref{fig:pooled_plenty_learn},\ref{fig:pooled_plenty_learnS}).
This is reasonable, as when resource depletion is removed it will reduce to the two-state model, which showed this behaviour.
However, when resource depletion is more severe, the second effect dominates and the mutant learns slower than wild-type (see \autoref{fig:pooled_scarce}\ref{fig:pooled_scarce_learn},\ref{fig:pooled_scarce_learnS}), which matches what is seen in the experiment.

@@

Gain-decrease pre-training will shift the distribution to the potentiated site, where there are more synapses available for depression and resources are more plentiful, for both the mutant and wild type.
This means that the pre-trained animal will learn faster then the untrained one for both mutant and wild-type (see \autoref{fig:pooled_plenty}\ref{fig:pooled_plenty_learnS} or \autoref{fig:pooled_scarce}\ref{fig:pooled_scarce_learnS}).
This differs from what is seen experimentally, where the pre-trained wild-type learns slower than the untrained one.

@@

With shorter pre-training, we see that it speeds up learning in both wild-type and mutant (see \autoref{fig:cascade_short}\ref{fig:cascade_short_learnS}), whereas experimentally this only happens for the mutant.
This is due to the fact that pre-training results in more synapses being potentiated, and thus ready for depression, but does not push them far enough down the cascade for the lower transition probabilities to slow down learning.

We can see from \autoref{fig:cascade_long}\ref{fig:cascade_long_learnS} that longer pre-training pushes the synapses further down the cascade, slow in down learning for the wild-type.
This effect is weaker for the mutant, as the equilibrium distribution is not as heavily concentrated at the end (see \autoref{fig:cascade_long}\ref{fig:cascade_long_eq}\ref{fig:cascade_long_eq_KO}).

>>>>

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:multiWTnopre}
  \Phi = -\frac{2\Delta fq}{M},
\end{equation}
%
where it's worth remembering that $\Delta f<0$.
With pre-training:
%
\begin{equation}\label{eq:multiWTpre}
\begin{aligned}
  \Phi &= 16(\Delta f)^2q \, \frac{(1-2\Delta f)^{M/2-1} (1+2\Delta f)^{M/2-1}}
          {(1-2\Delta f)^M - (1+2\Delta f)^M} \\
       &= -\frac{4\Delta fq}{M} + \CO(\Delta f)^2.
\end{aligned}
\end{equation}
%
So, we see that pre-training will speed up learning when $\Delta f$ is small (in absolute value), as seen in \autoref{fig:multistate_weak}\ref{fig:multistate_weak_learnS} (black curves).
On the other hand, if $\Delta f$ is close to $-\half$, pre-training will initially slow down learning a lot due to the factor of $(1+2\Delta f)^{M/2-1}$, as seen in \autoref{fig:multistate_strong}\ref{fig:multistate_strong_learnS}.

Intuitively, the flux depends on the slope of the distribution at the centre of the chain (with an offset due to the difference between $f\pot$ and $f\dep$).
Pre-training has two effects: it produces a slope in the right direction (see \autoref{fig:multistate_weak}\ref{fig:multistate_weak_eq}, red trace), but it also reduces the distribution at the centre (see \autoref{fig:multistate_strong}\ref{fig:multistate_strong_eq}, red trace).
For small $\Delta f$, the first effect is stronger and learning speeds up.
For larger $\Delta f$, the second effect wins and learning slows down.

@@

There are two effects here as well.
Smaller $\beta$ will increase the probability of crossing the centre of the chain, speeding up learning, but it will also concentrate probability at the ends of the chain, depleting the centre and slowing down learning (see \autoref{fig:multistate_strong}\ref{fig:multistate_strong_eq}\ref{fig:multistate_strong_eq_KO}, blue trace).
The first effect goes like $1/\beta$, whereas the second goes like $\beta^{M/2}$ and will be more significant for smaller $\beta$ or in a longer chain.

@@

However, this is only the initial, instantaneous rate of change.
Let's look at \autoref{fig:multistate_strong}\ref{fig:multistate_strong_learnS}, for which $\Delta f < \Delta f^*(1,M)$, so that pre-training initially slows down learning for both wild-type and mutant.
We see that the pre-trained curve can rapidly catch up with the un-pre-trained one, due to the mass of probability concentrated at the end of the chain (see \autoref{fig:multistate_strong}\ref{fig:multistate_strong_eq}, red trace) drifting to the centre.
This happens sooner for the mutant than the wild-type, due to the stronger depressing transitions.
This means that there is an intermediate range of time-scales over which pre-training does slow down learning in the wild-type but speed it up in the mutant, even outside the desired range of values for $\Delta f$.
As we see in \autoref{fig:multistate_med}\ref{fig:multistate_med_learnS} (black curves), even when we are in this range, the period of time for which pre-training impairs learning in the wild-type can be very small.

In conclusion, if we choose $\beta<\beta^*(M)$, $\Delta f < \Delta f^*(1,M)$ and look at intermediate time-scales, we will see that the mutant learns slower than wild-type without pre-training, and that pre-training speeds up learning in the mutant but slows it down in the wild-type.
However, in this case pre-training proceeds much faster in the mutant than wild-type (see \autoref{fig:multistate_med}\ref{fig:multistate_med_learn} and \autoref{fig:multistate_strong}\ref{fig:multistate_strong_learn}, lower curves), which is \emph{not} seen in the experiment.

@@

Finally, we note that, in the case of medium or strong pre-training, we see positive curvature in the learning curve after pre-training for a longer time in wild-type than in the mutant (see \autoref{fig:multistate_med}\ref{fig:multistate_med_learnS} and \autoref{fig:multistate_strong}\ref{fig:multistate_strong_learnS}, dashed curves).
This is due to the mass of probability at the right end of the chain (see \autoref{fig:multistate_med}\ref{fig:multistate_med_eq} and \autoref{fig:multistate_strong}\ref{fig:multistate_strong_eq}, red trace) that is unable to contribute to change in synaptic weight until it has drifted to the centre.
This will last for a shorter time in the mutant, due to the enhanced transitions.
This will not be a feature of any of the other models that we study here.
Now, the curvature can be changed by the nonlinear relation between synaptic weight and VOR gain, but both curves would be affected in the same way.
The difference in these two curvatures is therefore a good feature of this model.

@@

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:binWTnopre}
  \Phi = -\Delta f q,
\end{equation}
%
where, once again, it's worth remembering that $\Delta f<0$.
With pre-training:
%
\begin{equation}\label{eq:binWTpre}
\begin{aligned}
  \Phi &= -2\Delta f q.
\end{aligned}
\end{equation}
%
So, we see that pre-training will always speed up learning, unlike what is seen in the experiment.
This can be seen in \autoref{fig:binary}\ref{fig:binary_learnS} (black curves).

Now, consider the mutant, for which $q\pot=\beta q\dep=q$, $\beta<1$.
Without pre-training:
%
\begin{equation}\label{eq:binKOnopre}
  \Phi = -\frac{2\Delta f q}{1+\beta},
\end{equation}
%
which is always larger that \eqref{eq:binWTnopre}, unlike what is seen in the experiment.
This is also seen in \autoref{fig:binary}\ref{fig:binary_learn},\ref{fig:binary_learnS} (solid curves).
As discussed below \eqref{eq:multiKOnopre}, the smaller values of has two effects.
Smaller $\beta$ will increase the probability of depression, speeding up learning, but it will also decrease the probability of being ready for depression, slowing down learning.
The second effect is much smaller for the very short chain considered here.

@@

Now, consider the mutant, for which $q\pot=\beta q\dep=q$, $\beta<1$.
Without pre-training:
%
\begin{equation}\label{eq:multiLinKOnopre}
  \Phi = -2\Delta f q\,\frac{1-\beta^{M-1}}{1-\beta^M}.
\end{equation}
%
This decreases monotonically in the interval $\beta\in[0,1]$, therefore this will always be greater than \eqref{eq:multiLinWTnopre}.
This means that the mutant will initially learn faster than the wild-type.
However, as the wild-type will eventually catch up, and this can happen very quickly, as seen in \autoref{fig:multistate_lin}\ref{fig:multistate_lin_learn},\ref{fig:multistate_lin_learnS} (solid curves).

With pre-training:
%
\begin{equation}\label{eq:multiLinKOpre}
\begin{aligned}
  \Phi &= -4\Delta f q \, \frac{(1+2\Delta f)^{M-1} - \beta^{M-1}(1-2\Delta f)^{M-1}}
          {(1+2\Delta f)^M - \beta^M(1-2\Delta f)^M} \\
       &= -4\Delta f q\,\frac{1-\beta^{M-1}}{1-\beta^M} + \CO(\Delta f)^2.
\end{aligned}
\end{equation}
%
The ratio of this to \eqref{eq:multiLinKOnopre} takes its minimum value at the lowest end of the interval $\Delta f \in \brk{-\half,0}$, where it takes the value
%
\begin{equation}\label{eq:multiLinprenopre}
  \frac{\Phi_{\text{with pre}}}{\Phi_{\text{without pre}}} = \frac{1-\beta^M}{\beta-\beta^M}
   \longrightarrow \frac{M}{M-1} \quad \text{as} \quad \beta\to1,
\end{equation}
%
which is always greater than 1.
Therefore, pre-training will always enhance learning, for both mutant and wild-type, as seen in \autoref{fig:multistate_lin}\ref{fig:multistate_lin_learnS}.

@@

If we compare gain-increase learning in the mutant to wild-type, there are two effects: the increased transition rates speed up learning, but the equilibrium distribution is shifted to the depressed side (compare \autoref{fig:pooled_plenty}\ref{fig:pooled_plenty_eq}: \ref{fig:pooled_plenty_eq_WT} and \ref{fig:pooled_plenty_eq_KO} or \autoref{fig:pooled_scarce}\ref{fig:pooled_scarce_eq}: \ref{fig:pooled_scarce_eq_WT} and \ref{fig:pooled_scarce_eq_KO}, blue traces), where there are fewer synapses available for depression and resources are depleted.
When resource depletion is less severe, the first effect dominates and the mutant learns faster than wild-type (see \autoref{fig:pooled_plenty}\ref{fig:pooled_plenty_learn},\ref{fig:pooled_plenty_learnS}, solid curves).
This is reasonable, as when resource depletion is removed it will reduce to the two-state model, which showed this behaviour.
However, when resource depletion is more severe, the second effect dominates and the mutant learns slower than wild-type (see \autoref{fig:pooled_scarce}\ref{fig:pooled_scarce_learn},\ref{fig:pooled_scarce_learnS}, solid curves), which matches what is seen in the experiment.

@@

Gain-decrease pre-training will shift the distribution to the potentiated site, where there are more synapses available for depression and resources are more plentiful, for both the mutant and wild type (see \autoref{fig:pooled_scarce}\ref{fig:pooled_scarce_eq}, red traces).
This means that the pre-trained animal will learn faster then the untrained one for both mutant and wild-type (see \autoref{fig:pooled_plenty}\ref{fig:pooled_plenty_learnS} or \autoref{fig:pooled_scarce}\ref{fig:pooled_scarce_learnS}).
This differs from what is seen experimentally, where the pre-trained wild-type learns slower than the untrained one.

@@

With shorter pre-training, we see that it speeds up learning in both wild-type and mutant (see \autoref{fig:cascade_short}\ref{fig:cascade_short_learnS}), whereas experimentally this only happens for the mutant.
This is due to the fact that pre-training results in more synapses being potentiated, and thus ready for depression, but does not push them far enough down the cascade for the lower transition probabilities to slow down learning (see \autoref{fig:cascade_short}\ref{fig:cascade_short_pr}\ref{fig:cascade_short_pr_WT_pre}).

We can see from \autoref{fig:cascade_long}\ref{fig:cascade_long_pr}\ref{fig:cascade_long_pr_WT_pre} that longer pre-training pushes the synapses further down the cascade, slowing down learning for the wild-type.
This effect is weaker for the mutant, as the equilibrium distribution is not as heavily concentrated at the end (see \autoref{fig:cascade_long}\ref{fig:cascade_long_eq}\ref{fig:cascade_long_eq_KO}).
